\chapter{Combinatorial}

\section{The Twelvefold Way}
	Counts the \# of functions $f: N\to K, |N|=n, |K|=k$. The elements in $N$ and $K$ can be distinguishable or indistinguishable,
	while $f$ can be injective (one-to-one) of surjective (onto).
	\import{twelve.tex}
	Here, $S(n,k)$ is the Stirling number of the second kind, and $p(n,k)$ is the partition number.

\section{Permutations}
	\subsection{Factorial}
		\import{factorial.tex}
		\kactlimport{IntPerm.h}

	\subsection{Cycles}
		Let $g_S(n)$ be the number of $n$-permutations whose cycle lengths all belong to the set $S$. Then
		$$\sum_{n=0} ^\infty g_S(n) \frac{x^n}{n!} = \exp\left(\sum_{n\in S} \frac{x^n} {n} \right)$$

	\subsection{Derangements}
		Permutations of a set such that none of the elements appear in their original position.
		\[ \mkern-2mu D(n) = (n-1)(D(n-1)+D(n-2)) = n D(n-1)+(-1)^n = \left\lfloor\frac{n!}{e}\right\rceil \]
		\kactlimport{Derangements.h}

	\subsection{Involutions}
		An involution is a permutation with maximum cycle length 2, and it is its own inverse.
		\[ a(n)=a(n-1)+(n-1)a(n-2) \]
		\[ a(0)=a(1)=1 \]
		1,1,2,4,10,26,76,232,764,2620,9496,35696,140152

	\subsection{Burnside's lemma}
		Given a group $G$ of symmetries and a set $X$, the number of elements of $X$ \emph{up to symmetry} equals
		 \[ {\frac {1}{|G|}}\sum _{{g\in G}}|X^{g}|, \]
		 where $X^{g}$ are the elements fixed by $g$ ($g.x = x$).

		 If $f(n)$ counts ``configurations'' (of some sort) of length $n$, we can ignore rotational symmetry using $G = \mathbb Z_n$ to get
		 \[ g(n) = \frac 1 n \sum_{k=0}^{n-1}{f(\text{gcd}(n, k))} = \frac 1 n \sum_{k|n}{f(k)\phi(n/k)}. \]

\section{Partitions and subsets}
	\subsection{Partition function}
		Number of ways of writing $n$ as a sum of positive integers, disregarding the order of the summands.
		\[ p(n,k)=p(n-1,k-1)+p(n-k,k)\]
		\[ p(0,0)=p(1,n)=p(n,n)=p(n,n-1)=1\]
		For partitions with any number of parts, $p(n)$ obeys
		\[ p(0) = 1,\ p(n) = \sum_{k \in \mathbb Z \setminus \{0\}}{(-1)^{k+1} p(n - k(3k-1) / 2)} \]
		\[ p(n) \sim 0.145 / n \cdot \exp(2.56 \sqrt{n}) \]

		\begin{center}
		\begin{tabular}{c|c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c@{\ }c}
			$n$    & 0 & 1 & 2 & 3 & 4 & 5 & 6  & 7  & 8  & 9  & 20  & 50  & 100 \\ \hline
			$p(n)$ & 1 & 1 & 2 & 3 & 5 & 7 & 11 & 15 & 22 & 30 & 627 & $\mathtt{\sim}$2e5 & $\mathtt{\sim}$2e8 \\
		\end{tabular}
		\end{center}

	\subsection{Lucas' Theorem}
		Let $n,m$ be non-negative integers and $p$ a prime. Write $n=n_kp^k+...+n_1p+n_0$ and $m=m_kp^k+...+m_1p+m_0$. Then $\binom{n}{m} \equiv \prod_{i=0}^k\binom{n_i}{m_i} \pmod{p}$.

	\subsection{Binomials}
		\kactlimport{multinomial.h}

\section{General purpose numbers}
	\subsection{Bernoulli numbers}
		EGF of Bernoulli numbers is $B(t)=\frac{t}{e^t-1}$ (FFT-able).
		$B[0,\ldots] = [1, -\frac{1}{2}, \frac{1}{6}, 0, -\frac{1}{30}, 0, \frac{1}{42}, \ldots]$

		Sums of powers:
		\small
		\[ \sum_{i=1}^n n^m = \frac{1}{m+1} \sum_{k=0}^m \binom{m+1}{k} B_k \cdot (n+1)^{m+1-k} \]
		\normalsize

		Euler-Maclaurin formula for infinite sums:
		\small
		\[ \sum_{i=m}^{\infty} f(i) = \int_m^\infty f(x) dx - \sum_{k=1}^\infty \frac{B_k}{k!}f^{(k-1)}(m) \]
		\[ \approx \int_{m}^\infty f(x)dx + \frac{f(m)}{2} - \frac{f'(m)}{12} + \frac{f'''(m)}{720} + O(f^{(5)}(m)) \]
		\normalsize

	\subsection{Stirling numbers of the first kind}
		Number of permutations on $n$ items with $k$ cycles.
		\begin{align*}
			&c(n,k) = c(n-1,k-1) + (n-1) c(n-1,k)\\ 
			&c(0,0) = 1,c(0,n)=c(n,0)=0 \\
			&\textstyle \sum_{k=0}^n c(n,k)x^k = x(x+1) \dots (x+n-1)
		\end{align*}
		$c(8,k) = 8, 0, 5040, 13068, 13132, 6769, 1960, 322, 28, 1$ \\
		$c(n,2) = 0, 0, 1, 3, 11, 50, 274, 1764, 13068, 109584, \dots$

	\subsection{Eulerian numbers}
		Number of permutations $\pi \in S_n$ in which exactly $k$ elements are greater than the previous element. $k$ $j$:s s.t. $\pi(j)>\pi(j+1)$, $k+1$ $j$:s s.t. $\pi(j)\geq j$, $k$ $j$:s s.t. $\pi(j)>j$.
		$$E(n,k) = (n-k)E(n-1,k-1) + (k+1)E(n-1,k)$$
		$$E(n,0) = E(n,n-1) = 1$$
		$$E(n,k) = \sum_{j=0}^k(-1)^j\binom{n+1}{j}(k+1-j)^n$$

	\subsection{Stirling numbers of the second kind}
		Partitions of $n$ distinct elements into exactly $k$ groups.
		$$S(n,k) = S(n-1,k-1) + k S(n-1,k)$$
		$$S(n,1) = S(n,n) = 1$$
		$$S(n,k) = \frac{1}{k!}\sum_{j=0}^k (-1)^{k-j}\binom{k}{j}j^n$$

	\subsection{Bell numbers}
		Total number of partitions of $n$ distinct elements. $B(n) =$
		$1, 1, 2, 5, 15, 52, 203, 877, 4140, 21147, \dots$. For $p$ prime,
		\[ B(p^m+n)\equiv mB(n)+B(n+1) \pmod{p} \]

	\subsection{Labeled unrooted trees}
		\# on $n$ vertices: $n^{n-2}$ \\
		\# on $k$ existing trees of size $n_i$: $n_1n_2\cdots n_k n^{k-2}$ \\
		\# with degrees $d_i$: $(n-2)! / ((d_1-1)! \cdots (d_n-1)!)$

	\subsection{Catalan numbers}
		\[ C_n=\frac{1}{n+1}\binom{2n}{n}= \binom{2n}{n}-\binom{2n}{n+1} = \frac{(2n)!}{(n+1)!n!} \]
		\[ C_0=1,\ C_{n+1} = \frac{2(2n+1)}{n+2}C_n,\ C_{n+1}=\sum C_iC_{n-i} \]
		${C_n = 1, 1, 2, 5, 14, 42, 132, 429, 1430, 4862, 16796, 58786, \dots}$
		\begin{itemize}[noitemsep]
			\item sub-diagonal monotone paths in an $n\times n$ grid.
			\item strings with $n$ pairs of parenthesis, correctly nested.
			\item binary trees with with $n+1$ leaves (0 or 2 children).
			\item ordered trees with $n+1$ vertices.
			\item ways a convex polygon with $n+2$ sides can be cut into triangles by connecting vertices with straight lines.
			\item permutations of $[n]$ with no 3-term increasing subseq.
		\end{itemize}

\section{Young Tableaux}

	Let a \textbf{Young diagram} have shape $\lambda=(\lambda_1\geq \cdots\geq \lambda_k),$ where $\lambda_i$ equals the number of cells in the $i$-th (left-justified) row from the top. A \textbf{Young tableau} of shape $\lambda$ is a filling of the $n=\sum \lambda_i$ cells with a permutation of $1\ldots n$ such that each row and  column is increasing.

	\textbf{Hook-Length Formula}: For the cell in position $(i,j)$, let $h_\lambda(i,j)=\left|\{(I,J)|i\le I, j\le J, (I=i\text{ or }J=j)\}\right|.$ The number of Young tableaux of shape $\lambda$ is equal to
$f^\lambda \ =\ \frac{n!}{\prod h_\lambda (i,j)}.$

	\textbf{Schensted's Algorithm}: converts a permutation $\sigma$ of length $n$ into a pair of Young Tableaux $(S(\sigma),T(\sigma))$ of the same shape. When inserting $x=\sigma_i$,

	\begin{enumerate}
	\item Add $x$ to the first row of $S$ by inserting $x$ in place of the largest $y$ with $x<y$. If $y$ doesn't exist, push $x$ to the end of the row, set the value of $T$ at that position to be $i$, and stop.
	\item Add $y$ to the second row using the same rule, keep repeating as necessary.
	\end{enumerate}

	All pairs $(S(\sigma),T(\sigma))$ of the same shape correspond to a unique $\sigma$, so
	$n!=\sum (f^{\lambda})^2.$ Also, $S(\sigma^R)=S(\sigma)^T.$

	Let $d_k(\sigma),a_k(\sigma)$ be the lengths of the longest subseqs which are a union of $k$ decreasing/ascending subseqs, respectively. Then $a_k(\sigma)=\sum_{i=1}^k\lambda_i, d_k(\sigma)=\sum_{i=1}^k\lambda^*_i,$ where $\lambda^*_i$ is size of the $i$-th column. 

\section{General prupose theorems}

	\subsection{Identities}
	Vandermonde Convolution: $\binom{m+n}{r} = \sum_{k=0}^r\binom{m}{k}\cdot\binom{r}{n-k}$.

	Hockey Stick: $\binom{n+1}{r+1} = \sum_{i=r}^n\binom{i}{r}$.

	\subsection{Cycle Lemma}
	Any sequence of $m X$'s and $n Y$'s, where $m > n$ has exactly $m -n$ cyclic permutations which are dominating,
	and $m - kn$ which are $k$-dominating. To find them, arrange sequence in a circle and repeatedly remove
	adjacent pairs $XY$. The remaining $X$'s were each the start of a dominating permutation.

	\subsection{Sprague Grundy theorem}
	Every impartial game is equivalent to a nimber. Nimbers are de-fined inductively as $*0 = \{\}, *1 = {*0}, *2 = {*0, *1}, *(n+1) = *n \cup {n}$, and
	corresponds to a heap of size $n$. The formula for adding positions is 

	\[ S + S' = {S + s' | s' \in S'} \cup {s + S' | s \in S} \]
	\[ a + b = a \oplus b + 2(a\&b) \]

	Define minimum exclusion $M : \phi(N) \rightarrow N$ by $M(S) = $ the least non-negative integer not in S.
	Let $C = (M(A) \oplus B) \cup (M(B) \oplus A)$. Then $M(C) = M(A) \oplus M(B)$.
	Define $SG(S) = M (\{SG(s) | s \in S\})$.
	$SG(Nim_k) = k$ by strong induction. Game is losing iff $SG(S) = 0$. Theorem: $SG(A + B) = SG(A) \oplus SG(B)$.

	\subsection{Partisan Game}
	Can define the negative of a game by interchanging $L$ and $R$'s possible moves.
	Define $G = 0$ if first player loses. $G = H$ if $G + (-H) = 0$.
	A cold game is one which moving only hurts players. In this case we never have G fuzzy 0, so G is
	representable as an integer, thus calculable by DP.

	\subsection{Matrices for operators}
	Matrices for xor, and, and or are: $\frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 1 \\ 1 & -1\end{bmatrix}, \begin{bmatrix} 0 & 1 \\ 1 & 1\end{bmatrix}$
	with inverses: $\begin{bmatrix} -1 & 1 \\ 1 & 0\end{bmatrix}, \begin{bmatrix} 1 & 1 \\ 1 & 0\end{bmatrix}$.

	\subsection{Prufer sequences}
	The set of labeled trees on $n$ vertices corresponds bijectively to the set of Prufer sequences of length $n-2$.
	To convert a tree into a Prufer sequence, repeatedly remove the leaf with the smallest label, and write down its neighbor.
	To convert sequence to tree, first set the degree of each vertex to $n_v + 1$, where $n_v$ is the number of times
	the vertex appears in the sequence. Then for each $i$, find lowest $j$ with degree $1$, add edge $a_i, j$, and decrease
	the degrees of $a_i$ and $j$ by $1$. After this, two nodes of degree 1 remain - connect them.

	This can be used to calculate number of labeled trees in a complete bipartite graph - $l^{r-1}\cdot r^{l-1}$.

	\subsection{Tournament Graphs}
	There exists a Hamiltonian path on any tournament graphs - use induction to find. Cycle if strongly connected. TFAE:
	$T$ is 1. transitive. 2. is strict total ordering. 3. is acyclic. 4. has no cycle of length $3$. 5. The outdegrees are $\{0, 1, \cdots, n-1\}$.
	 6. has exactly one Hamiltonian path.

	\subsection{Landau's theorem}
	A sequence of numbers is called a score sequence if for each subset $S$, sum of 
	numbers in $S$ is at least $\binom{|S|}{2}$ and sum of all numbers is $\binom{n}{2}$.

	This score sequence represents the outdegrees of a vertex in a tournament graph.

	\subsection{Dilworth's / Hall's / Mirsky's theorem}
	Maximum antichain has same size as minimum chain decomposition.

	Maximum chain size has same size as minimum antichain decomposition.

	To compute size, model as bipartite graph with two copies of vertices - $v_in$ and $v_out$.
	Distinct representatives can be chosen for a family of sets $S$ iff every subfamily $W$ of $S$ has at least $|W|$
	elements in their union. E.g. Left side of bipartite graph can be fully matched iff each subset has sufficient ”degree”.

	\subsection{Laplacian Matrix and Kirchoff’s Theorem}
	Laplacian matrix is defined as $L = D - A$, where $D$ is the degree matrix (diagonal), and $A$ the adjacency matrix.

	Kirchoff’s Theorem states that the number of spanning trees in a graph is any cofactor of the Laplacian.

	To calculate that, remove the first row and column and calculate the determinant of the remaining matrix.

\section{Other}
	\kactlimport{NimProduct.h}
	% \kactlimport{SchreierSims.h}
	\kactlimport{DeBruijnSeq.h}
	\kactlimport{MatroidIntersect.h}
